A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-11-08 latex/main.tex
Updates from Sally

@@

The most comprehensive analysis on recombination in SARS-CoV-2 to 
date has been performed using the combination of UShER and 
RIPPLES~\cite{Turakhia2022}. 
% NOTE: not citing RIVET here as there is no real analysis
% in the paper itself, it's just focused on the tool
RIPPLES works by taking an UShER phylogeny and systematically breaking up
sequences subtended by long branches into multiple segments 
that are placed independently.
A recombination event at a node is flagged if
the overall parsimony of these multiple placements
sufficiently improves over
the original single-parent origin in the phylogeny.
This approach has much in common with the LS HMM
used in sc2ts, but with one important distinction.
While sc2ts detects and incorporates recombination
in ``real-time'' as sequences are added to the ARG in daily
batches, RIPPLES operates on a complete phylogeny,
attempting to detect recombination post-hoc among millions of nodes. 
Like any phylogenetic reconstruction method
that ignores recombination, UShER must force placement of recombinant
sequences in places that minimize the discrepancy across the entire sequence,
which may not be near any of the true parents. 
These permanent misplacements distort the phylogeny~\cite{Schierup2000consequences}
and subsequent tree reorganisations to maximise overall parsimony~\cite{DeMaio2023} 
may break up the artefactual long branches used to detect recombination.

We compared the performance of sc2ts recombinant detection 
against RIPPLES run on the Viridian UShER phylogeny,
subset to the 2,475,418 samples shared with the ARG (STAR Methods).
To enable exact comparison, we omitted the QC filtering steps from both
methods, and initially set the RIPPLES ``parsimony improvement'' 
parameter to $p=4$ (rather than the more lenient default of 3), 
to match the equivalent $k=4$ setting used in sc2ts (STAR Methods). 
Under this setting,
RIPPLES reported 1174 recombination events of which 
601 were validated by 3SEQ (51\%; STAR Methods).
% JK: There's no point in making a meal of the 3SEQ validation rate,
% This is notably lower than the equivalent 
% concordance rate of 57\% among the 855 recombination events reported by sc2ts.
We then compared
the specific events reported by both methods by aligning them
using sets of descendant samples (STAR Methods),
giving 468 events shared between RIPPLES and sc2ts (Figure~\ref{fig:ripples_sc2ts_events}).
The 706 recombination events output by RIPPLES that do not 
correspond to a recombination event in the sc2ts ARG can largely
be explained by a combination of the choice of sites 
used for inference, and the strict time ordering imposed
by sc2ts (Document~\ref{sec:detailed_ripples}).
The 387 events detected by sc2ts and not by UShER+RIPPLES
would require a detailed dissection of the UShER tree
to explain, which is beyond the scope of this work.
However, the set does contain some well characterised events 
associated with the Pango X lineages.
Table~\ref{tab:ripples_x_lineages} shows the events detected 
by UShER+RIPPLES associated with samples assigned to Pango X
lineages, showing that at $p=4$
only 9 of the Pango X lineages are present and
with unequivocal recombinant origin for 6 
(contrast with 16 Type I events in Table~\ref{tab:pango_x_lineages}).
We therefore reran RIPPLES with $p=3$, resulting in 4111 events overall,
and increasing the intersection of events with sc2ts to 595 (70\%;
Figure~\ref{fig:ripples_sc2ts_events}).
At this more lenient threshold, 23 of the 44 Pango X lineages
are identified as descendants of RIPPLES recombinant events,
with unequivocal origins for 10 lineages, and
7 lineages associated with two competing events 
(Document~\ref{sec:detailed_ripples}).
XB and XAJ are among the lineages detected by UShER+RIPPLES,
strengthening evidence that these are false negatives 
for sc2ts (Table~\ref{tab:pango_x_lineages}).

The choice of the $p$ threshold has a large effect 
on the number of recombination events detected by RIPPLES, and 
previous studies have used the default value of 
$3$\cite{Turakhia2022,Smith2023tracking}.
For example, when applied to an UShER phylogeny of
$\sim$1.6 million genomes sampled up to May, 2021,
RIPPLES reported 589 unique recombination events with 43,104 descendant
samples--corresponding to $\sim$2.7\% of all samples of recombinant origin.
Over the same time interval, the sc2ts ARG ($k=4$) identifies
14 recombination events passing QC with
172 descendant samples among 501,235 samples (0.034\%),
a 40- to 80-fold reduction.
The sc2ts ARG identifies more of the well-characterised Pango X lineages
at a more stringent threshold, and therefore provides us with a 
conservative lower bound
on the frequency of detectable recombination events in SARS-CoV-2.

\section*{DISCUSSION}
There are two primary contributions of this work. The first is an
ARG for 2.48 million SARS-CoV-2 genomes.
Through multiple layers of validation against results from the 
literature, we have demonstrated that this ARG synthesises
the evolutionary features of genetic inheritance, point mutations,
deletions and recombination in a single cohesive structure.
The ARG is freely available and easily accessible using the 
tskit software ecosystem, a rich and efficient platform for 
recombination-aware analyses widely used in population genomics.
This combination of a deeply validated pandemic-scale ARG with 
a mature and highly capable software platform forms a unique resource.
Previous landmark ARGs~\cite{Wohns2022,Anderson2023genes}
have helped spur the development of ARG-aware methods in human 
genomics~\cite{Mejia2025using,Grundler2025geographic,
Lehmann2025args,Terhorst2025accelerated}.
We hope that the ARG provided here will similarly
enable a range of recombination-aware analyses of SARS-CoV-2, and 
catalyse a wave of methodological developments integrating
recombination into pathogen genomics.
For instance, annotating the ARG with experimentally derived
fitness estimates for mutations can help elucidate selection
pressures on different recombinants;
analysing the genomic positions of mutations occurring on long
branches (which persist after accounting for recombination) can
illuminate the evolution of saltational variants.
We envisage the ARG as a long-term stable resource---building on the 
high-quality, freely available reassemblies of
the Viridian dataset~\cite{Hunt2024}---to facilitate the study 
of the pandemic phase of SARS-CoV-2 in an open and reproducible manner.

The other primary contribution of this work is the method used to 
infer the ARG, sc2ts. The ability to infer an ARG over millions of whole 
genomes is a step change in scale over existing ARG 
inference methods for pathogen data\cite{Didelot2010inference,%
Vaughan2017inferring,%
Muller2020bayesian,%
Ignatieva2022,%
Muller2022bayesian}.
To ensure these innovations are well-founded, we have paid significant 
attention to methodological detail, carefully validating the ARG
(and consequently the method) against known results. 
Sc2ts is currently specialised to take advantage of unique features 
of the SARS-CoV-2 dataset, but the methods may be readily adapted
to other pathogens. The LS HMM is the most important element of 
sc2ts, efficiently finding parsimonious placements for samples 
among all possible recombinant solutions. The current 
parametrisation---emphasising interpretability---is simplistic,
and could readily be extended to account
for mutation rate variation along the genome~\cite{Haddox2025}
or transition-to-transversion rate bias~\cite{Bloom2023,Simmonds2024}.
Our analysis of deletions shows that valuable phylogenetic signal
is lost when they are ignored, and an important avenue for 
future work is to explore ways of incorporating them into the HMM.
Another methodological innovation is the use of 
the tsdate method~\cite{Wohns2022,Pope2025} to estimate internal node dates
in the ARG, illustrating the reuse of methods initially developed 
for population genetic applications. While results are encouraging, 
dating of phylogenies is a complex task~\cite{Kumar2016advances}
and there is certainly room for improvement.

Further contributions of this work arise from 
the detailed study of the sc2ts ARG,
providing insights into the evolution of SARS-CoV-2
and the COVID-19 pandemic. 
First, we clarify the relationships among
recombinant Pango X lineages, finding examples of
clustering of named lineages under a single shared recombination event.
Second, we used the ARG to study breakpoint intervals 
for recombination events, adding a novel temporal dimension
and demonstrating a simple relationship with the 
size of the interval in base pairs derived from theory.
Thirdly, we demonstrate a clear relationship between
genetic diversity and the rate of detectable recombination,
with the best predictor of the number of recombination
events over time being the product of global case numbers
and lineage diversity.
Finally, and perhaps most fundamentally, through our 
conservative parameter choices and real-time handling of 
recombination events, we provide a stringent lower
bound on the amount of recombination that occurred during
the pandemic phase of SARS-CoV-2.
These results underscore the utility of the ARG 
and provide a template for future analyses in which 
% JK: Going out on a limb here, please do change if 
% we don't like it!!!
recombination is fully integrated rather than treated
as a nuisance.

@@

The collection date of a sample
is an important piece of information for sc2ts.
We used the Viridian consensus date (``Date\_tree``),
based on information from COG-UK, GISAID, and ENA/SRA as our source.
Samples which have missing or incomplete dates (e.g., 2020-01) were omitted,
as were the $\sim$40,000 samples
which had a collection date of December 31, 2020.
As the sample count for this date is a major outlier
for this period of the pandemic,
it seems likely that this batch is highly enriched
for incorrect collection dates.
Manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer)
confirmed that some of the samples were not collected in 2020.

@@

Alignments read from the VCF Zarr file were first preprocessed
to mask any non-nucleotide characters (A, C, G, or T) as missing data.
% https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian_config.toml
Then, we filtered any samples that had missing data at more than 500 sites,
excluding the 100 sites identified as problematic
(this was set to 10,000 sites for the initial period up to 2020-03-01
where sampling density is low).

\subsubsection*{Li and Stephens model}

@@

We use the Viterbi algorithm implemented in tsinfer~\cite{Kelleher2019}
to efficiently find copying paths under this model. We updated this 
implementation to include a ``likelihood threshold'', which allows 
all possible paths through the HMM with likelihood less than a given
value to be considered equally probable. This approach leads to major
gains in likelihood compression, saving memory and reducing computation
time. We then use this likelihood thresholding approach to iteratively
process the samples for a given daily batch, first finding all exact 
matches, then all with 1 mismatch, etc. By setting the appropriate
likelihood threshold for the HMM, we are guaranteed that any HMM
solutions found at this threshold are correct, and can then run the HMM again
with a lower threshold for the remaining samples. Close matches
are found very quickly using this approach, and it is only
samples that are more distant from the current ARG that require
significant processing time and memory. This is reflected in 
the computational resources reported in 
Figures~\ref{fig:inference_resources},\ref{fig:samples_per_day}.

@@

The reversion push operation inspects a newly added node to see
if any of its mutations are ``immediate reversions''; that is,
are reversions of a mutation that occurred on the new node's
immediate parent. We increase the overall parsimony of the
inference by ``pushing in'' a new node which descends from the
original parent, and carries all its mutations except those
causing the reversions on the newly added node.
52,336 nodes were added to the ARG during primary inference by
reversion pushing.

\subsubsection*{Filtering time travellers}
Time-travelling samples---those in which the recorded collection date differs
substantially from their true collection date---pose significant problems
for sc2ts, and are present at a significant frequency. For example, of the 
43 Scorpio~\cite{Colquhoun2023scorpio}
designations present in the Viridian dataset studied, 32 have 
samples present with sampling dates in 2020, including lineages that do not
arise until late 2022. Inserting these time-travellers into the ARG 
at the claimed sampling date causes major issues, and they must therefore be 
filtered out. We use the simple approach of filtering out samples that 
exceed a given ``HMM cost'', i.e., requiring too many mutations or
recombination breakpoints to plausibly represent the steady accumulation
of diversity expected with dense sampling. 
The HMM cost for the Viterbi solution for a sample under the LS model
is $k$ times the number of recombination events plus the number of 
mismatches, where $k$ is the recombination penalty (here, $k=4$).
After some experimentation, we set the HMM cost threshold to 7, 
disallowing samples that require 8 or more mutations (or, e.g., 1 recombination
and 4 mutations, or 2 recombinations) from immediate inclusion (see next
section) in the ARG. This approach has the benefit of filtering
samples enriched for errors as well as time-travellers.

@@

\subsubsection{XL}
\label{sec:dd_xl}
All the XL samples descend from a well supported recombination node,
with three child samples which are identical to it.
The left parent is a sample (ERR7989424), and
the right parent is a reconstructed internal node
assigned to BA.2.5 with a large number of descendant samples.

@@

\begin{table}
\centering
\caption{
Sc2ts detection results for Pango X lineages not added to the ARG.
The number of samples in Viridian v04 dataset
assigned to these Pango X lineages,
the number of samples which passed sequence-level quality control, and
the number of recombination breakpoints
in the Viterbi solutions of the samples
are shown.
}
\label{tab:extra_hmm_results}
\begin{tabular}{crrr}
\toprule
Pango & Samples in Viridian & Samples passing QC & Breakpoints \\
\midrule
XBJ & 2 & 1 & 1\\
XBP & 8 & 2 & 1\\
XBS & 19 & 15 & 1\\
XBW & 1 & 1 & 1\\
XCA & 11 & 4 & 1\\
XAK & 2 & 2 & 1\\   % Multiple breakpoint Pango X
XAY & 15 & 11 & 6\\ % Multiple breakpoint Pango X
XBC & 36 & 21 & 3\\ % Multiple breakpoint Pango X
XBL & 2 & 2 & 0\\   % Multiple breakpoint Pango X
XBT & 1 & 1 & 2\\   % Multiple breakpoint Pango X
\bottomrule
\end{tabular}
\end{table}

>>>>

\newcommand{\red}[1]{\textcolor{red}{#1}}

@@

The most comprehensive analysis on recombination in SARS-CoV-2 to 
date has been performed using the combination of UShER and 
RIPPLES~\cite{Turakhia2022}. 
% NOTE: not citing RIVET here as there is no real analysis
% in the paper itself, it's just focused on the tool
RIPPLES works by taking an UShER phylogeny and systematically breaking up
sequences subtended by long branches into multiple segments 
that are placed independently.
A recombination event at a node is flagged if
the overall parsimony of these multiple placements
sufficiently improves over
the original single-parent origin in the phylogeny.
This approach has much in common with the LS HMM
used in sc2ts, but with one important distinction.
While sc2ts detects and incorporates recombination
in ``real-time'' as sequences are added to the ARG in daily
batches, RIPPLES operates on a complete phylogeny,
attempting to detect recombination post-hoc among millions of nodes. 
Like any phylogenetic reconstruction method
that ignores recombination, UShER must force placement of recombinant
sequences in places that minimise the discrepancy across the entire sequence,
which may not be near any of the true parents. 
These permanent misplacements distort the phylogeny~\cite{Schierup2000consequences},
and subsequent tree reorganisations to maximise overall parsimony~\cite{DeMaio2023} 
may break up the artefactually long branches used to detect recombination.

We compared the performance of sc2ts to detect recombination 
against RIPPLES run on the Viridian UShER phylogeny,
subset to the 2,475,418 samples shared with the ARG (STAR Methods).
To enable exact comparison, we omitted the QC filtering steps from both
methods, and initially set the RIPPLES ``parsimony improvement'' 
parameter to $p=4$ (rather than the more lenient default of 3), 
to match the equivalent $k=4$ setting used in sc2ts (STAR Methods). 
Under this setting,
RIPPLES reported 1174 recombination events of which 
601 were validated by 3SEQ (51\%; STAR Methods).
% JK: There's no point in making a meal of the 3SEQ validation rate,
% This is notably lower than the equivalent 
% concordance rate of 57\% among the 855 recombination events reported by sc2ts.
We then compared
the specific events reported by both methods by aligning them
using sets of descendant samples (STAR Methods),
giving 468 events shared between RIPPLES and sc2ts (Figure~\ref{fig:ripples_sc2ts_events}).
The 706 recombination events detected by RIPPLES that do not 
correspond to a recombination event in the sc2ts ARG can largely
be explained by a combination 
of the strict time ordering imposed by sc2ts
and the differing sets of problematic sites masked
by sc2ts and UShER
(Document~\ref{sec:detailed_ripples}).
The 387 events detected by sc2ts and not by UShER+RIPPLES
 include some of the well characterised Pango X lineages.
Table~\ref{tab:ripples_x_lineages} shows the events detected 
by UShER+RIPPLES associated with samples assigned to Pango X
lineages, showing that at $p=4$
only 9 of the Pango X lineages are present and
with unequivocal recombinant origin for 6 
(contrast with 16 Type I events in Table~\ref{tab:pango_x_lineages}).
We therefore reran RIPPLES with $p=3$, resulting in 4111 events overall
and increasing the intersection of events with sc2ts to 595 
(Figure~\ref{fig:ripples_sc2ts_events}).
At this more lenient threshold, 23 of the 44 Pango X lineages
are identified as descendants of RIPPLES recombinant events,
with unequivocal origins for 10 lineages, and
7 lineages associated with two competing events 
(Document~\ref{sec:detailed_ripples}).
Thus, a more lenient threshold ($p=3$) was needed for UShER+RIPPLES to detect a similar fraction of Pango X lineages as sct2ts with $k=4$, resulting in more false positives. That said, XB and XAJ are among the lineages detected by UShER+RIPPLES, even at $p=4$,
strengthening evidence that these are false negatives 
for sc2ts (Table~\ref{tab:pango_x_lineages}).

The choice of the $p$ threshold has a large effect 
on the number of recombination events detected by RIPPLES, and 
previous studies have used the default value of 
$3$\cite{Turakhia2022,Smith2023tracking}.
For example, when applied to an UShER phylogeny of
$\sim$1.6 million genomes sampled up to May, 2021,
RIPPLES reported 589 unique recombination events with 43,104 descendant
samples---corresponding to $\sim$2.7\% of all samples of recombinant origin.
Over the same time interval, the sc2ts ARG ($k=4$) identifies
14 recombination events passing QC with
172 descendant samples among 501,235 samples (0.034\%),
a 40- to 80-fold reduction.
The sc2ts ARG identifies more of the well-characterised Pango X lineages
at a more stringent threshold, and therefore provides us with a 
conservative lower bound
on the frequency of detectable recombination events in SARS-CoV-2.

\section*{DISCUSSION}
There are two primary contributions of this work. The first is an
ARG for 2.48 million SARS-CoV-2 genomes.
Through multiple layers of validation against results from the 
literature, we have demonstrated that this ARG synthesises
the evolutionary features of genetic inheritance, point mutations,
deletions and recombination in a single cohesive structure.
The ARG is freely available and easily accessible using the 
tskit software ecosystem, a rich and efficient platform for 
recombination-aware analyses widely used in population genomics.
This combination of a deeply validated pandemic-scale ARG with 
a mature and highly capable software platform forms a unique resource.
Previous landmark ARGs~\cite{Wohns2022,Anderson2023genes}
have helped spur the development of ARG-aware methods in human 
genomics~\cite{Mejia2025using,Grundler2025geographic,
Lehmann2025args,Terhorst2025accelerated}.
We hope that the ARG provided here will similarly
enable a range of recombination-aware analyses of SARS-CoV-2 and 
catalyse a wave of methodological developments integrating
recombination into pathogen genomics.
For instance, annotating the ARG with experimentally derived
fitness estimates for mutations can help elucidate selection
pressures on different recombinants;
saltational variants can also be detected by analysing the rate and position of mutations along long
branches (which persist after accounting for recombination).
We envisage the ARG as a long-term stable resource---building on the 
high-quality, freely available reassemblies of
the Viridian dataset~\cite{Hunt2024}---to facilitate the study 
of the pandemic phase of SARS-CoV-2 in an open and reproducible manner.

The other primary contribution of this work is the method used to 
infer the ARG, sc2ts. The ability to infer an ARG over millions of whole 
genomes is a step change in scale over existing ARG 
inference methods for pathogen data\cite{Didelot2010inference,%
Vaughan2017inferring,%
Muller2020bayesian,%
Ignatieva2022,%
Muller2022bayesian}.
To ensure these innovations are well-founded, we have paid significant 
attention to methodological detail, carefully validating the ARG
(and consequently the method) against known results. 
Sc2ts is currently specialised to take advantage of unique features 
of the SARS-CoV-2 dataset, but the methods may be readily adapted
to other pathogens. The LS HMM is the most important element of 
sc2ts, efficiently finding parsimonious placements for samples 
among all possible recombinant solutions. The current 
parametrisation---emphasising interpretability---is simplistic
and could readily be extended to account
for mutation rate variation along the genome~\cite{Haddox2025}
or transition-to-transversion rate bias~\cite{Bloom2023,Simmonds2024}.
Our analysis of deletions shows that valuable phylogenetic signal
is lost when they are ignored, and an important avenue for 
future work is to explore ways of incorporating them directly into the HMM.
Another methodological innovation is the use of 
the tsdate method~\cite{Wohns2022,Pope2025} to estimate internal node dates
in the ARG, illustrating the reuse of methods initially developed 
for population genetic applications. While results are encouraging, 
dating of phylogenies is a complex task~\cite{Kumar2016advances},
and there is certainly room for improvement.

Further contributions of this work arise from 
the detailed study of the sc2ts ARG,
providing insights into the evolution of SARS-CoV-2
and the COVID-19 pandemic. 
First, we clarify the relationships among
recombinant Pango X lineages, finding examples of
clustering of named lineages under a single shared recombination event.
Second, we used the ARG to study breakpoint intervals 
for recombination events, showing that the relationship between the 
size of the interval in base pairs and the divergence time of the parental lineages matches theoretical expectations.
Thirdly, we demonstrate a clear relationship between
genetic diversity and the rate of detectable recombination,
with the best predictor of the number of recombination
events over time being the product of global case numbers
and lineage diversity.
Finally, and perhaps most fundamentally, through our 
conservative parameter choices and real-time handling of 
recombination events, we provide a stringent lower
bound on the amount of recombination that occurred during
the pandemic phase of SARS-CoV-2.
These results underscore the utility of the ARG, and 
the efficient and reproducible pipelines throughout this work 
provide a template for future analyses in which recombination
is accounted for by default.

@@

The collection date of a sample
is an important piece of information for sc2ts.
We used the Viridian consensus date (``Date\_tree``),
based on information from COG-UK, GISAID, and ENA/SRA as our source.
Samples that have missing or incomplete dates (e.g., 2020-01) were omitted. 
Samples with a collection date of December 31, 2020
were also omitted,
as the number of sample on this date ($\sim$40,000) was a major outlier
for this period of the pandemic and is likely enriched
for incorrect collection dates.
Manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer)
confirmed that some of these samples were not collected in 2020.

@@

Alignments read from the VCF Zarr file were first preprocessed
to mask any non-nucleotide characters (A, C, G, or T) as missing data.
% https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian_config.toml
Then, excluding the 100 sites identified as problematic, we filtered any samples that had missing data at more than 500 sites (this was set to 10,000 sites for the initial period up to 2020-03-01
when sampling density was low).
% SO: Moved clause about 100 sites to the beginning so that it was clearer that 10,000 is in comparison to 500 not 100.
\subsubsection*{Li and Stephens model}

@@

We use the efficient Viterbi algorithm
(which finds the most likely sequence of hidden states in a HMM) 
implemented in tsinfer~\cite{Kelleher2019}
to find copying paths under this model. We updated this 
implementation to include a ``likelihood threshold'', which allows 
all possible paths through the HMM with likelihood less than a given
value to be considered equally improbable. This approach leads to major
gains in likelihood compression, saving memory and reducing computation
time. We then use this likelihood thresholding approach to iteratively
process the samples for a given daily batch, first finding all exact 
matches, then all with 1 mismatch, etc. By setting the appropriate
likelihood threshold for the HMM, we are guaranteed that any HMM
solutions found at this threshold are correct, and can then run the HMM again
with a lower threshold for the remaining samples. Close matches
are found very quickly using this approach, and it is only
samples that are more distant from the current ARG that require
significant processing time and memory. This is reflected in 
the computational resources reported in 
Figures~\ref{fig:inference_resources},\ref{fig:samples_per_day}.

@@

The reversion push operation inspects a newly added node to see
if any of its mutations are ``immediate reversions''; that is,
are reversions of a mutation that occurred on the new node's
immediate parent. We increase the overall parsimony of the
inference by ``pushing in'' a new node which descends from the
original parent and carries all its mutations except those
causing the reversions on the newly added node.
Reversion pushing led to the addition of 52,336 nodes during primary inference.

\subsubsection*{Filtering time travellers}
Time-travelling samples---those in which the recorded collection date differs
substantially from their true collection date---pose significant problems
for sc2ts and are present at a significant frequency. For example, of the 
43 Scorpio~\cite{Colquhoun2023scorpio}
designations present in the Viridian dataset studied, 32 have 
samples present with sampling dates in 2020, including lineages that do not
arise until late 2022. Inserting these time-travellers into the ARG 
at the claimed sampling date causes major artefacts, and they must therefore be 
filtered out. We use the simple approach of filtering out samples that 
exceed a given ``HMM cost'', i.e., requiring too many mutations or
recombination breakpoints to plausibly represent the steady accumulation
of diversity expected with dense sampling. 
The HMM cost for the Viterbi solution for a sample under the LS model
is $k$ times the number of recombination events plus the number of 
mismatches, where $k$ is the recombination penalty (here, $k=4$).
After some experimentation, we set the HMM cost threshold to 7, 
disallowing samples that require 8 or more mutations (or, e.g., 1 recombination
and 4 mutations, or 2 recombinations) from immediate inclusion (see next
section) in the ARG. This approach has the benefit of filtering
samples enriched for errors as well as time-travellers.

@@

\subsubsection{XL}
\label{sec:dd_xl}
All the XL samples descend from a well supported recombination node,
with three child samples, which are identical to it.
The left parent is a sample (ERR7989424), and
the right parent is a reconstructed internal node
assigned to BA.2.5 with a large number of descendant samples.

@@

\begin{table}
\centering
\caption{
Sc2ts detection results for Pango X lineages not added to the ARG.
The number of samples in Viridian v04 dataset
assigned to these Pango X lineages,
the number of samples that passed sequence-level quality control, and
the number of recombination breakpoints
in the Viterbi solutions of the samples
are shown.
}
\label{tab:extra_hmm_results}
\begin{tabular}{crrr}
\toprule
Pango & Samples in Viridian & Samples passing QC & Breakpoints \\
\midrule
XBJ & 2 & 1 & 1\\
XBP & 8 & 2 & 1\\
XBS & 19 & 15 & 1\\
XBW & 1 & 1 & 1\\
XCA & 11 & 4 & 1\\
XAK & 2 & 2 & 1\\   % Multiple breakpoint Pango X
XAY & 15 & 11 & 6\\ % Multiple breakpoint Pango X
XBC & 36 & 21 & 3\\ % Multiple breakpoint Pango X
XBL & 2 & 2 & 0\\   % Multiple breakpoint Pango X
XBT & 1 & 1 & 2\\   % Multiple breakpoint Pango X
\bottomrule
\end{tabular}
\end{table}
